---
title: "Problem Set_04"
author: "Helen Lin (30776142)"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---
### Problem set_04 “Fine-scale phylogenetic architecture”
#### Learning objectives: 

* Gain experience estimating diversity within a hypothetical microbial community

#### Part 1: Description and enumeration
Obtain a collection of "microbial" cells from "seawater". The cells were concentrated from different depth intervals by a marine microbiologist travelling along the Line-P transect in the northeast subarctic Pacific Ocean off the coast of Vancouver Island British Columbia. 

Sort out and identify different microbial "species" based on shared properties or traits. Record your data in this Rmarkdown using the example data as a guide.

Once you have defined your binning criteria, separate the cells using the sampling bags provided. These operational taxonomic units (OTUs) will be considered separate "species". This problem set is based on content available at [What is Biodiversity](http://cnx.org/content/m12179/latest/).

For example, load in the packages you will use.
```{r message=FALSE}
#To make tables
library(kableExtra)
library(knitr)
#To manipulate and plot data
library(tidyverse)
```

Then load in the data. You should use a similar format to record your community data.
```{r}
example_data1 = data.frame(
  number = c(1,2,3),
  name = c("lion", "tiger", "bear"),
  characteristics = c("brown cat", "striped cat", "not a cat"),
  occurences = c(2, 4, 1)
)
```

Finally, use these data to create a table.
```{r}
example_data1 %>% 
  kable("html") %>%
  kable_styling(bootstrap_options = "striped", font_size = 10, full_width = F)
```
	
For your sample:

```{r}
#This is the population
library(kableExtra)
library(knitr)
library(tidyverse)

diversity_candy = data.frame(
  number = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14),
  name = c("Skittles", "Vines", "Mikes & Ikes", "Bricks", "Gummy Bears", "M&M", "Sour Bears", "Sour Fruit", "Hershey Kisses", "Jujubes", "Sour Bottle", "Sour Swirl", "Wine Candy", "Sour Hexa"),
  characteristics = c("chewy sour candy", "red string candy", "rod shaped fruit candy", "brick shaped candy", "gummy bear candy", "circular chocolate candy", "sour bear candy", "sour fruit shaped candy", "cone shaped chocolate candy", "hard gummy candy", "sour bottle-shaped candy", "sour swirl candy", "wine candy", "sour 6-legged candy"),
  occurences = c(187, 14, 174, 18, 101, 241, 3, 2, 16, 24, 3, 3, 9, 6))

kable(diversity_candy, "html") %>% 
  kable_styling(bootstrap_options = "striped", font_size = 10, full_width = F)
```

* Construct a table listing each species, its distinguishing characteristics, the name you have given it, and the number of occurrences of the species in the collection.

```{r}
diversity_candy_sample = data.frame(
  number = c(1,2,3,4,5,6,7,8,9,10,11),
  name = c("Skittles", "Vines", "Mikes & Ikes", "Bricks", "Gummy Bears", "M&M", "Hershey Kisses", "Jujubes", "Sour Swirl", "Wine Candy", "Sour Hexa"),
  characteristics = c("chewy sour candy", "red string candy", "rod shaped fruit candy", "brick shaped candy", "gummy bear candy", "circular chocolate candy", "cone shaped chocolate candy", "hard gummy candy", "sour swirl candy", "wine candy", "sour 6-legged candy"),
  occurences = c(49, 8, 40, 3, 23, 44, 2, 3, 1, 1, 2))

kable(diversity_candy_sample, "html") %>% 
  kable_styling(bootstrap_options = "striped", font_size = 10, full_width = F)
```

* Ask yourself if your collection of microbial cells from seawater represents the actual diversity of microorganisms inhabiting waters along the Line-P transect. Were the majority of different species sampled or were many missed? 
 
#### Part 2: Collector's curve
To help answer the questions raised in Part 1, you will conduct a simple but informative analysis that is a standard practice in biodiversity surveys. This analysis involves constructing a collector's curve that plots the cumulative number of species observed along the y-axis and the cumulative number of individuals classified along the x-axis. This curve is an increasing function with a slope that will decrease as more individuals are classified and as fewer species remain to be identified. If sampling stops while the curve is still rapidly increasing then this indicates that sampling is incomplete and many species remain undetected. Alternatively, if the slope of the curve reaches zero (flattens out), sampling is likely more than adequate.

To construct the curve for your samples, choose a cell within the collection at random. This will be your first data point, such that X = 1 and Y = 1. Next, move consistently in any direction to a new cell and record whether it is different from the first. In this step X = 2, but Y may remain 1 or change to 2 if the individual represents a new species. Repeat this process until you have proceeded through all cells in your collection.

For example, we load in these data.
```{r}
example_data2 = data.frame(
  x = c(1,2,3,4,5,6,7,8,9,10),
  y = c(1,2,3,4,4,5,5,5,6,6)
)
```
My own data:

```{r}
candy_sample_collector_ = data.frame(x = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62,63,64,65,66,67,68,69,70,71,72,73,74,75,76,77,78,79,80,81,82,83,84,85,86,87,88,89,90,91,92,93,94,95,96,97,98,99,100,101,102,103,104,105,106,107,108,109,110,111,112,113,114,115,116,117,118,119,120,121,122,123,124,125,126,127,128,129,130,131,132,133,134,135,136,137,138,139,140,141,142,143,144,145,146,147,148,149,150,151,152,153,154,155,156,157,158,159,160,161,162,163,164,165,166,167,168,169,170,171,172,173,174,175),
  y = c(1,2,3,3,3,4,4,5,6,7,8,8,9,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11,11)
)

```
And then create a plot. We will use a scatterplot (geom_point) to plot the raw data and then add a smoother to see the overall trend of the data.
```{r}
ggplot(example_data2, aes(x=x, y=y)) +
  geom_point() +
  geom_smooth() +
  labs(x="Cumulative number of individuals classified", y="Cumulative number of species observed")
```

My plot:

```{r}
ggplot(candy_sample_collector_, aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method="loess") +
  labs(x="Cumulative number of individuals classified", y="Cumulative number of species observed")

```
For your sample:

* Create a collector's curve.
* Does the curve flatten out? If so, after how many individual cells have been collected?
* What can you conclude from the shape of your collector's curve as to your depth of sampling?

#### Part 3: Diversity estimates (alpha diversity)
Using the table from Part 1, calculate species diversity using the following indices or metrics.

##### Diversity: Simpson Reciprocal Index

$\frac{1}{D}$ where $D = \sum p_i^2$

$p_i$ = the fractional abundance of the $i^{th}$ species

For example, using the example data 1 with 3 species with 2, 4, and 1 individuals each, D =

```{r}
species1 = 2/(2+4+1)
species2 = 4/(2+4+1)
species3 = 1/(2+4+1)

1 / (species1^2 + species2^2 + species3^2)
```

The higher the value is, the greater the diversity. The maximum value is the number of species in the sample, which occurs when all species contain an equal number of individuals. Because the index reflects the number of species present (richness) and the relative proportions of each species with a community (evenness), this metric is a diveristy metric. Consider that a community can have the same number of species (equal richness) but manifest a skewed distribution in the proportion of each species (unequal evenness), which would result in different diveristy values.

* What is the Simpson Reciprocal Index for your sample?

```{r}
species1 = 49/175
species2 = 8/175
species3 = 40/175
species4 = 3/175
species5 = 23/175
species6 = 44/175
species7 = 2/175
species8 = 3/175
species9 = 1/175
species10 = 1/175
species11 = 2/175

1/(species1^2 + species2^2 + species3^2 + species4^2 + species5^2 + species6^2 + species7^2 + species8^2 + species9^2 + species10^2 + species11^2)

## Ans = 4.66986886
```

* What is the Simpson Reciprocal Index for your original total community?

```{r}

species1 = 187/801
species2 = 14/801
species3 = 174/801
species4 = 18/801
species5 = 101/801
species6 = 241/801
species7 = 3/801
species8 = 2/801
species9 = 16/801
species10 = 24/801
species11 = 3/801
species12 = 3/801
species13 = 9/801
species14 = 6/801

1/(species1^2 + species2^2 + species3^2 + species4^2 + species5^2 + species6^2 + species7^2 + species8^2 + species9^2 + species10^2 + species11^2 + species12^2 + species13^2 + species14^2)

## Answer is 4.75164967
```

##### Richness: Chao1 richness estimator
Another way to calculate diversity is to estimate the number of species that are present in a sample based on the empirical data to give an upper boundary of the richness of a sample. Here, we use the Chao1 richness estimator. 

$S_{chao1} = S_{obs} + \frac{a^2}{2b})$

$S_{obs}$ = total number of species observed
a = species observed once
b = species observed twice or more

So for our previous example community of 3 species with 2, 4, and 1 individuals each, $S_{chao1}$ =

```{r}
3 + 1^2/(2*2)
```

* What is the chao1 estimate for your sample?

```{r}
11 + 6^2/(2*5)

##Ans = 14.6
```

#### Part 4: Alpha-diversity functions in R
We've been doing the above calculations by hand, which is a very good exercise to aid in understanding the math behind these estimates. Not surprisingly, these same calculations can be done with R functions. Since we just have a species table, we will use the `vegan` package. You will need to install this package if you have not done so previously. 

```{r message=FALSE}
library(vegan)
```

First, we must remove the unnecesary data columns and transpose the data so that `vegan` reads it as a species table with species as columns and rows as samples (of which you only have 1).
```{r}
example_data1_diversity = 
  example_data1 %>% 
  select(name, occurences) %>% 
  spread(name, occurences) 

example_data1_diversity
```

Then we can calculate the Simpson Reciprocal Index using the `diversity` function.
```{r}
diversity(example_data1_diversity, index="invsimpson")
```

And we can calculate the Chao1 richness estimator (and others by default) with the the `specpool` function for extrapolated species richness. This function rounds to the nearest whole number so the value will be slightly different that what you've calculated above.
```{r}
specpool(example_data1_diversity)
```

In Project 1, you will also see functions for calculating alpha-diversity in the `phyloseq` package since we will be working with data in that form.

For your sample:

* What is the Simpson Reciprocal Index using the R function?

```{r}
diversity_candy_diversity =
  diversity_candy %>% 
  select(name, occurences) %>% 
  spread(name, occurences)

diversity_candy_diversity

diversity(diversity_candy_diversity, index="invsimpson")
## [1] 4.75165
```
* What is the chao1 estimate using the R function?

```{r}
specpool(diversity_candy_diversity)
## chao = 14
```
    + Verify that these values match your previous calculations.

#### Part 5: Concluding activity
If you are stuck on some of these final questions, reading the [Kunin et al. 2010](https://www.ncbi.nlm.nih.gov/pubmed/19725865) and [Lundin et al. 2012](https://www.ncbi.nlm.nih.gov/pubmed/23760801) papers may provide helpful insights. 

* How does the measure of diversity depend on the definition of species in your samples?
* Can you think of alternative ways to cluster or bin your data that might change the observed number of species?
* How might different sequencing technologies influence observed diversity in a sample?